Data Exploration: IUCN and AquaMaps

### 2015-10-13 data_explore.Rmd
fixed html formatting, fixed column names from rerun of spp_list

# Data files

iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells.csv'), col_types = 'idiii', progress = TRUE)

@@

### Species Map Function ###
# This function takes a single species scientific name as input, then grabs all 
# occurrence cells and associated Aquamaps probability and/or IUCN proportional area
# per cell
get_spp_map <- function(species){
  spp_id <- spp_list %>%
    filter(sciname == species & is.na(parent_sid)) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    unique()
  iucn_spp_map <- iucn_spp_cells %>%
    filter(iucn_sid == spp_id$iucn_sid) %>%
    dplyr::select(-parent_sid, -subpop_sid, loiczid = iucn_loiczid) %>%
    group_by(loiczid, iucn_sid) %>%        
    summarize(iucn_area = max(iucn_area))
  ### The group_by() and summarize() are to collapse cells with overlapped polygons, fixing the duplicate 'by' issue
  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid) %>%
    rename(loiczid = am_loiczid)
  spp_map <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid') %>%
    as.data.frame()

@@

  print(i)

@@

  #get just the aquamaps cells
#   am_spp_map = spp_map%>%
#                   filter(!is.na(am_prob))%>%
#                 dplyr::select(loiczid,am_sid,am_prob)%>% #add in am_prob when we want to look at probability
#                   unique()

#   if(TRUE %in% duplicated(am_spp_map$loiczid)){
#                   am_spp_map = am_spp_map%>%
#                                 dplyr::select(loiczid,am_sid)%>%
#                                   unique()}
#   
  #rasterize aquamaps map
#   r_am_spp  <-  subs(loiczid_raster, 
#                      am_spp_map[ , c('loiczid', 'am_sid')], 
#                      by = 'loiczid', 
#                      which = 'am_sid', 
#                      subsWithNA = TRUE)
#   
#   spp_list[i,]$am_area = area(r_am_spp,na.rm=T)%>%
#               cellStats(.,stat='sum')
#

  #get the iucn cells
#   iucn_map <- spp_map %>%
#     mutate(iucn_pres = iucn_area > 0)%>%
#       dplyr::select(loiczid,iucn_area,am_sid,am_prob,iucn_pres)%>%
#         unique()

@@

  #this is incorporated because there are some duplications for loiczids with IUCN cells. I'm guessing this is because of the parent/subpopulations?

#    if(TRUE %in% duplicated(iucn_map$loiczid))next()
# #   
#  rasterize IUCN map
#   r_iucn_spp <- subs(loiczid_raster, 
#                      iucn_map[ , c('loiczid', 'iucn_pres')], 
#                     by = 'loiczid', 
#                      which = 'iucn_pres', 
#                      subsWithNA = TRUE)
# #   
#   spp_list[i,]$iucn_area = area(r_iucn_spp,na.rm=T)%>%
#                 cellStats(.,stat='sum')

  # Get total area (IUCN and AM combined)
  allCells <- spp_map%>%
                dplyr::select(loiczid)%>%
                  mutate(value=1)%>%
                  unique()

@@

  # get percent overlap of map

#   #iucn map cells
#   cells_iucn<-iucn_map%>%
#               filter(iucn_pres==TRUE)%>%
#               .$loiczid
#   cells_am <- am_spp_map%>%
#                 filter(!is.na(am_sid))%>%
#                 .$loiczid
#   
#   cells_total = unique(spp_map$loiczid)
#   cells_overlap = intersect(cells_iucn,cells_am)
#   
#    r_iucn_spp <- subs(loiczid_raster, 
#                       iucn_map[ , c('loiczid', 'iucn_pres')], 
#                      by = 'loiczid', 
#                       which = 'iucn_pres', 
#                       subsWithNA = TRUE)
#   
  #percent overlap
#   perc = (length(cells_overlap)/length(cells_total))*100
#   spp_list[i,]$perc = perc
#   
#   am_only = setdiff(cells_am,cells_iucn)
#   iucn_only = setdiff(cells_iucn,cells_am)
#   
#   perc_am = (length(am_only)/length(cells_total))*100
#   perc_iucn = (length(iucn_only)/length(cells_total))*100
#   
#   spp_list[i,]$am_perc = perc_am
#   spp_list[i,]$iucn_perc = perc_iucn
# 
#   spp_list[i,]$sm_range = ifelse(length(cells_am)<length(cells_iucn),'AM','IUCN') #what dataset has the smaller range
#   spp_list[i,]$sm_perc = length(cells_overlap)/min(length(cells_am),length(cells_iucn))*100 #this is the total percent of smaller range encompassed in the larger range
# cat(perc,perc_am,perc_iucn)
#

@@

write.csv(spp_list,file='spp_list_w_area_percent.csv')
```

@@

spp_list = read.csv('spp_list_w_area_percent.csv')

@@

dot = ggplot(df,aes(iucn_area,am_area,color=category))+ geom_point() + geom_abline()+
      labs(x='Area (km2) from IUCN dataset',y='Area (km2) from AquaMaps dataset',title='Comparing total area of species range maps')
ggsave('figures/CompareTotalArea_AM-IUCN_scatter.png',plot=dot,width=8, height=4, dpi=100)
print(dot)
```

@@

all_status <- ggplot(df,aes(iucn_log,am_log,color=category)) + 
  geom_point() + geom_abline() +
  labs(x='IUCN Area (log transformed)', y='AquaMaps Area (log transformed)',
        title= 'Comparing area of range maps\n (all area values are km2 log transformed)',fill='IUCN Category')
ggsave('figures/CompareTotalAreaLOG_AM-IUCN_scatter.png',plot=all_status,width=8,height=4,dpi=100)
print(all_status)
```

@@

groups <- ggplot(df, aes(iucn_log, am_log,color=category)) + geom_point() +facet_wrap(~spp_group) + 
        geom_abline() + labs(x='IUCN Area (log transformed)', y='AquaMaps Area (log transformed)',
                             title='Comparing differences in area across species groups', fill='IUCN Category')
ggsave('figures/CompareTotalAreaLOG_AM-IUCN_spgroups_scatter.png',plot=groups,width=8,height=6,dpi=100)
print(groups)
```
***

@@

  paste0(round(p1[1, ], 2), '% of the total area in this map is covered by both IUCN and AquaMaps')
  paste0(round(p1[2, ], 2), '% of the total area in this map is covered just by AquaMaps')
  paste0(round(p1[3, ], 2), '% of the total area in this map is covered just by IUCN')
```

@@

Split up by species - not sure if this is useful or not
```{r,echo=F}

@@

Looking at the percent of smaller species range that is within the larger range.

```{r pct small within large, echo=F}

@@

Exploring possible correlation between total species range area and percent overlap of smaller range within the larger:
``` {r pct overlap vs area, echo = FALSE}

@@

``` {r, echo = FALSE}

@@

Perent coverage as function of area, clipping to area below 100,000,000 km^2:
``` {r, echo = F}
# remove all area greater than 100000000

@@

Percent coverage as a function of log(area)
``` {r, echo = F}
# Log transform area

@@

``` {r examine regressions, echo = FALSE}

@@

Examining data broken into quartiles to see if smaller patterns emerge
``` {r examine quartiles, echo = FALSE}

@@

qtiles <- c(quantile(spp_list$total_area, 0.0),
            quantile(spp_list$total_area, .25),
            quantile(spp_list$total_area, .50),
            quantile(spp_list$total_area, .75),
            quantile(spp_list$total_area, 1.0))
qtile_mdl <- list()
for (i in 1:4) {
  area_clip <- spp_list %>%
    filter(total_area >= qtiles[i] & total_area < qtiles[i + 1])
  qtile_mdl[[i]] <- lm(total_area ~ sm_perc, area_clip)

@@

Looking at how many and what type (threat category) of species we are losing by setting a threshold of 0.4 on the aquamaps species.

```{r aquamaps probabilities, warning=F,message=F,cache.lazy=FALSE,cache=T,verbose=F,results=F,echo=F}

>>>>

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN species
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_both.csv'), col_types = 'cdddd', progress = TRUE)%>%as.data.frame()%>%filter(presence!=5)

@@

### Species Map Function ###
# This function takes a single species scientific name as input, then grabs all 
# occurrence cells and associated Aquamaps probability and/or IUCN proportional area
# per cell
get_spp_map <- function(species){
  spp_id <- spp_list %>%
    filter(sciname == species & is.na(parent_sid)) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    unique()
  iucn_spp_map <- iucn_spp_cells %>%
    filter(sciname == species) %>%
    group_by(loiczid) %>%        
    summarize(iucn_area = max(prop_area))
  ### The group_by() and summarize() are to collapse cells with overlapped polygons, fixing the duplicate 'by' issue
  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid) %>%
    rename(am_prob = prob)
  spp_map <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid')%>%
    as.data.frame()

@@

@@

###get just the aquamaps cells
   am_spp_map = spp_map%>%
                   filter(!is.na(am_prob))%>%
                    dplyr::select(loiczid,am_sid,am_prob)%>% #add in am_prob when we want to look at probability
                     unique()

  if(TRUE %in% duplicated(am_spp_map$loiczid)){
                  am_spp_map = am_spp_map%>%
                                dplyr::select(loiczid,am_sid)%>%
                                  unique()}

###  rasterize aquamaps map
  r_am_spp  <-  subs(loiczid_raster, 
                     am_spp_map[ , c('loiczid', 'am_sid')], 
                     by = 'loiczid', 
                     which = 'am_sid', 
                     subsWithNA = TRUE)

  spp_list[i,]$am_area = area(r_am_spp,na.rm=T)%>%
              cellStats(.,stat='sum')

###  get the iucn cells
  iucn_map <- spp_map %>%
    mutate(iucn_pres = iucn_area > 0)%>%
      dplyr::select(loiczid,iucn_area,am_sid,am_prob,iucn_pres)%>%
        unique()

@@

###this is incorporated because there are some duplications for loiczids with IUCN cells. I'm guessing this is because of the parent/subpopulations?

   if(TRUE %in% duplicated(iucn_map$loiczid))next()

### rasterize IUCN map
  r_iucn_spp <- subs(loiczid_raster, 
                     iucn_map[ , c('loiczid', 'iucn_pres')], 
                    by = 'loiczid', 
                     which = 'iucn_pres', 
                     subsWithNA = TRUE)

  spp_list[i,]$iucn_area = area(r_iucn_spp,na.rm=T)%>%
                cellStats(.,stat='sum')

### Get total area (IUCN and AM combined)
  allCells <- spp_map%>%
                dplyr::select(loiczid)%>%
                  mutate(value=1)%>%
                  unique()

@@

### get percent overlap of map

   #iucn map cells
  cells_iucn<-iucn_map%>%
              filter(iucn_pres==TRUE)%>%
              .$loiczid
  cells_am <- am_spp_map%>%
                filter(!is.na(am_sid))%>%
                .$loiczid

  cells_total = unique(spp_map$loiczid)
  cells_overlap = intersect(cells_iucn,cells_am)

  #I THINK THIS IS A DUPLICATE? r_iucn_spp raster created above... (JA 10/12/15)
#    r_iucn_spp <- subs(loiczid_raster, 
#                       iucn_map[ , c('loiczid', 'iucn_pres')], 
#                      by = 'loiczid', 
#                       which = 'iucn_pres', 
#                       subsWithNA = TRUE)
#   
  #percent overlap
  perc = (length(cells_overlap)/length(cells_total))*100
  spp_list[i,]$perc = perc

  am_only = setdiff(cells_am,cells_iucn)
  iucn_only = setdiff(cells_iucn,cells_am)

  perc_am = (length(am_only)/length(cells_total))*100
  perc_iucn = (length(iucn_only)/length(cells_total))*100

  spp_list[i,]$am_perc = perc_am
  spp_list[i,]$iucn_perc = perc_iucn

  spp_list[i,]$sm_range = ifelse(length(cells_am)<length(cells_iucn),'AM','IUCN') #what dataset has the smaller range
  spp_list[i,]$sm_perc = length(cells_overlap)/min(length(cells_am),length(cells_iucn))*100 #this is the total percent of smaller range encompassed in the larger range
cat(sprintf('%s: For species %s, %.2f%% of both datasets overlap, %.2f%% is just AquaMaps and %.2f%% is just IUCN\n',i,spp,perc,perc_am,perc_iucn))

@@

#remove non IUCN species and log total area (for use later on)
spp_list = spp_list%>%
            filter(spp_group != 'hagfishes' & spp_group != 'non-homalopsids')%>%
             mutate(log_total_area = log(total_area))

write.csv(spp_list,file='spp_list_w_area_percent_noextinct.csv')
```

@@

spp_list = read.csv('spp_list_w_area_percent_noextinct.csv')

@@

dot = ggplot(df,aes(iucn_area,am_area,color=iucn_category))+ geom_point() + geom_abline()+
      labs(x='Area (km2) from IUCN dataset',y='Area (km2) from AquaMaps dataset',title='Comparing total area of species range maps')
ggsave('figures/CompareTotalArea_AM-IUCN_scatter.png',plot=dot,width=8, height=4, dpi=100)
print(dot)
```

@@

all_status <- ggplot(df,aes(iucn_log,am_log,color=iucn_category)) + 
  geom_point() + geom_abline() +
  labs(x='IUCN Area (log transformed)', y='AquaMaps Area (log transformed)',
        title= 'Comparing area of range maps\n (all area values are km2 log transformed)',fill='IUCN Category')
ggsave('figures/CompareTotalAreaLOG_AM-IUCN_scatter.png',plot=all_status,width=8,height=4,dpi=100)
print(all_status)
```

@@

groups <- ggplot(df, aes(iucn_log, am_log,color=iucn_category)) + geom_point() +facet_wrap(~spp_group) + 
        geom_abline() + labs(x='IUCN Area (log transformed)', y='AquaMaps Area (log transformed)',
                             title='Comparing differences in area across species groups', fill='IUCN Category')

ggsave('figures/CompareTotalAreaLOG_AM-IUCN_spgroups_scatter.png',plot=groups,width=8,height=6,dpi=100)
print(groups)
```
***

@@

For `r spp`, `r round(pi[1], 2)`% of the total area is covered by both IUCN and AquaMaps, `r round(p1[2,],2)`% is covered just by AquaMaps and `r round(p1[3,], 2)`% is covered just by IUCN.

@@

Split up by species - not sure if this is useful or not
```{r,echo=F, message = F, warning = F}

@@

###(4) Looking at the percent of smaller species range that is within the larger range.

For each range

```{r pct small within large, echo=F, warning= F}

@@

Exploring possible correlation between total species range area and percent overlap of smaller range within the larger:
``` {r pct overlap vs area, echo = FALSE, warning = F, message=F}

@@

``` {r, echo = FALSE, warning = F, message=F}

@@

Perent coverage as function of area, clipping to area below 100,000,000 km^2:
``` {r, echo = F, warning = F, message = F}
# remove all area greater than 100000000

@@

Percent coverage as a function of log(area)
``` {r, echo = F, warning = F, message = F}
# Log transform area

@@

``` {r examine regressions, echo = FALSE, warning = F}

@@

###(5) Examining data broken into quartiles to see if smaller patterns emerge
``` {r examine quartiles, echo = FALSE, warning = F}

@@

qtiles <- c(quantile(spp_list$total_area, 0.0,na.rm=T),
            quantile(spp_list$total_area, .25,na.rm=T),
            quantile(spp_list$total_area, .50,na.rm=T),
            quantile(spp_list$total_area, .75,na.rm=T),
            quantile(spp_list$total_area, 1.0,na.rm=T))

qtile_mdl <- list()
for (i in 1:4) {
  area_clip <- spp_list %>%
    filter(total_area >= qtiles[i] & total_area < qtiles[i + 1])
  qtile_mdl[[i]] <- lm(total_area ~ sm_perc, area_clip)

@@

###(6) Looking at how many and what type (threat category) of species we are losing by setting a threshold of 0.4 on the aquamaps species.

```{r aquamaps probabilities, warning=F,message=F,cache.lazy=FALSE,cache=T,verbose=F,results=F,echo=F,include=F}
